---
title: "Exploring the New Coder Survey 2016"
author: "Muhsina Elampilakkattil"
date: "12 November 2018"
output: 
  html_document:
    toc: true
    fig_width: 6
    fig_height: 3
    code_folding: hide
---
<hr/>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Summary

Free Code Camp is an open source community where you learn to code and build projects for nonprofits.Free Code Camp surveyed more than 15,000 people in 2016 who started coding within the past 5 years.

The goal of this survey was to understand these people's motivations in learning to code, how they're learning to code, their demographics, and their socioeconomic background.

<hr/>
##Load Libraries

```{r, message= FALSE, warning= FALSE}
library(tidyverse)
library(RColorBrewer)
library(ggthemes)
library(gridExtra)
library(cowplot)

fillBlue= "#2071a3"
fillLBlue= "#c3d8e5"
fillSilver= "#C0C0C0"

survey <- read_csv("Datasets/2016-FCC-New-Coders-Survey-Data.csv")

```

##Who participated?
Majority of the respondents are 27 years old. The youngest new coder is 10 years old and the oldest one is 86 years.
```{r, warning= FALSE, fig.width= 10, fig.height= 9}
#Age distribution
plot_age<- survey %>%
  filter(!is.na(Age)) %>%
  ggplot(aes(x= Age)) +
  geom_histogram(binwidth= 2, fill= fillBlue) + 
geom_vline(aes(xintercept = mean(Age)), 
             linetype = "dashed", size = 0.6,
             color = "#FC4E07")+
  labs(title= "Age Distribution", x= "Age", y= "Count", size= 2) + 
geom_text(aes(x= mean(Age), y= -200, label=round(mean(Age),digits=2)),hjust=-0.2, size=2, color= "red") +
  theme_minimal(base_size= 8)  

#Gender Distribution
plot_gender<- survey %>%
  filter(!is.na(Gender)) %>%
  group_by(Gender) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  ungroup() %>%
  ggplot(aes(x= reorder(Gender, count), y= count)) +
  geom_bar(stat= "identity", fill= fillBlue) +
  geom_text(aes(label= paste0(round(pct,2),"%", sep="")),
      position= position_dodge(0), color= "black", size= 2) +
  labs(title= "Gender Distribution", x= "", y= "Count") +
  coord_flip() +
  theme_minimal(base_size= 8)

#How many months have you been programming for?
plot_exp <- survey %>%
  filter(!is.na(MonthsProgramming)) %>%
  ggplot(aes(x= MonthsProgramming))+
  geom_density(aes(y = ..count..), fill = "lightgray", color= fillBlue, alpha= 0.9) + scale_x_log10() +
  geom_vline(aes(xintercept = mean(MonthsProgramming)), 
             linetype = "dashed", size = 0.6,
             color = "#FC4E07") +
   labs(title= "Programming experience", x= "Months", y= "Count", size= 2) + 
  geom_text(aes(x= mean(MonthsProgramming), y= -500, label=round(mean(MonthsProgramming),digits=2)),hjust=-0.2, size=2, color= "red") +
  theme_minimal(base_size= 8)

#Hours Spend learning each week
plot_learning<- survey %>%
  filter(!is.na(HoursLearning)) %>%
  ggplot(aes(x= HoursLearning))+
  geom_density(aes(y = ..count..), fill = "lightgray", color= fillBlue, alpha= 0.9) + scale_x_log10() +
  geom_vline(aes(xintercept = mean(HoursLearning)), 
             linetype = "dashed", size = 0.6,
             color = "#FC4E07") +
   labs(title= "Hours learning per week", x= "hours", y= "Count", size= 2) + 
  geom_text(aes(x= mean(HoursLearning), y= -1000, label=round(mean(HoursLearning),digits=2)),hjust=-0.2, size=2, color= "red") +
  theme_minimal(base_size= 8)

#Which country are you a citizen of?
plot_country<- survey %>%
  filter(!is.na(CountryCitizen)) %>%
  group_by(CountryCitizen) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(10) %>%
  ungroup() %>%
  
  ggplot(aes(x= reorder(CountryCitizen, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
  geom_text(aes(label = paste0(round(pct,2),"%", sep="")),
      position= position_dodge(0), color= "black", size= 2) + 
  labs(title= "Country of citizenship", x= "") +
  coord_flip() + theme_minimal(base_size= 8)

#Which language do you speak at home with your family?
plot_lang<- survey %>%
  filter(!is.na(LanguageAtHome)) %>%
  group_by(LanguageAtHome) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(10) %>%
  ggplot(aes(x= reorder(LanguageAtHome, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
  geom_text(aes(label = paste0(round(pct,2),"%", sep="")),
      position= position_dodge(0), color= "black", size= 2) + 
  labs(title= "Language at home", x= "") +
  coord_flip() + theme_minimal(base_size= 8)

first_row<- plot_grid(plot_age, plot_gender, align= 'h', rel_widths= c(1,1))
second_row<- plot_grid(plot_exp, plot_country, align= 'h', rel_widths= c(1,1.2))
third_row<- plot_grid(plot_learning, plot_lang, align='h',  rel_widths= c(1,1))
plot_grid(first_row, second_row, third_row, ncol= 1, rel_heights= c(1,1,1))

#Summary
summary(survey$Age)

```

##Which Job Role are you most Interested in?
40% of the new coders answered "*Full-stack web developer" to the question.
```{r}
#Job roles interested
survey %>%
  filter(!is.na(JobRoleInterest)) %>%
  group_by(JobRoleInterest) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ,
        DS= ifelse(JobRoleInterest=="Data Scientist / Data Engineer", TRUE, FALSE)) %>%
  ggplot(aes(x= reorder(JobRoleInterest, count), y = count, fill= DS)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c(fillSilver, fillBlue)) +
  geom_text(aes(label = paste0(round(pct, digits= 1),"%", sep="")), position= position_dodge(0),hjust= 0, size=2) +
  labs(title= "Job role interest", x="", y="Count") +
  coord_flip() + 
  ylim(c(0,3000)) +
  theme_minimal(base_size= 8)+
  theme(legend.position = "none")
```


## Data Science/ Data Engineers
Lets dive deeper into the details of developing Data scientists or Data engineers.

```{r, fig.width= 10}
#Subsetting the data
DS<- filter(survey, JobRoleInterest == "Data Scientist / Data Engineer")

#Age distribution
DS_age<- DS %>%
  filter(!is.na(Age)) %>%
  ggplot(aes(x= Age)) +
  geom_histogram(binwidth= 0.02, fill= fillBlue) + 
  labs(title= "Age Distribution", x= "Age", y= "Count", size= 2) + 
  scale_x_log10() +
  theme_minimal(base_size= 8) 

#Gender Distribution
DS_gender<- DS %>%
  filter(!is.na(Gender)) %>%
  group_by(Gender) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  ungroup() %>%
  ggplot(aes(x= reorder(Gender, count), y= count, fill= count)) +
  geom_bar(stat= "identity") +
  geom_text(aes(label= paste0(round(pct,1),"%", sep="")),
      position= position_dodge(0), hjust= 1.5, color= "white", size= 3) +
  labs(title= "What is your gender?", x= "", y= "Count") +
  scale_fill_gradient(low= fillLBlue, high= fillBlue) +
  coord_flip() +
  theme_minimal(base_size= 8) +
    theme(legend.position = "none")

plot_grid(DS_gender, DS_age, align="h", rel_widths = c(1,1.2))
```

## Programming experience and learning style

```{r, fig.width= 10, warning= FALSE}
#How many months have you been programming for?
DS_exp <- DS %>%
  filter(!is.na(MonthsProgramming)) %>%
  ggplot(aes(x= MonthsProgramming))+
  geom_histogram(binwidth= 0.9, fill = fillBlue) +
  labs(title= "Programming experience", x= "Months", y= "Count", size= 2) + 
  scale_x_continuous(breaks= c(1, 8, 60, 360), labels= c("1", "8", "60", "360"), trans= "log2") +
  theme_minimal(base_size= 8)

#Hours Spend learning each week
DS_learning<- survey %>%
  filter(!is.na(HoursLearning)) %>%
  ggplot(aes(x= HoursLearning))+
  geom_histogram(binwidth= 0.5, fill = fillBlue) +
  labs(title= "Hours learning per week", x= "hours", y= "Count", size= 2) + 
  scale_x_continuous(breaks= c(5,10,20,40,80),labels= c("5","10","20","40","80"), trans="log2") +
  theme_minimal(base_size= 8)

plot_grid(DS_exp, DS_learning, align= "h")
```

### Job Preference

```{r}
#Job Preference
DS %>%
  group_by(JobPref) %>%
  filter(!is.na(JobPref)) %>%
  summarize(count = n()) %>%
  rbind(data.frame("JobPref"= c("start your own business", "freelance"), count= c(0,0))) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  ggplot(aes(x= reorder(JobPref, count), y = count, fill= count)) + 
  geom_bar(stat = "identity") + 
  scale_x_discrete(drop=FALSE) +
  labs(title="Job preference", x="", y= "Count") +
  geom_text(aes(label = paste0(round(pct, digits= 2),"%")), hjust= 1, color= "white", size= 3) + 
  coord_flip() +
  scale_fill_gradient(low= fillLBlue, high=fillBlue) +
  theme_minimal(base_size= 8) +
  theme(legend.position= "none")

#where would you prefer to work?
DS %>%
  filter(!is.na(JobWherePref)) %>%
  group_by(JobWherePref) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  ggplot(aes(x= reorder(JobWherePref, count), y = count, fill= count)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = paste0(round(pct, 1),"%")), color="white", hjust=1, size= 2.5) + 
  labs(title= "Where would you prefere to work?", x="", y="Count")+
  scale_fill_gradient(low=fillLBlue, high= fillBlue) +
  coord_flip()+
  theme_minimal(base_size= 8) +
  theme(legend.position= "none")

#JobApplyWhen
DS %>%
  filter(!is.na(JobApplyWhen)) %>%
  group_by(JobApplyWhen) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  ggplot(aes(x= reorder(JobApplyWhen, count), y = count, fill= count)) + 
  geom_bar(stat = "identity") + 
  labs(title= "Applying for Jobs?", x="",y="Count") +
  geom_text(aes(label = paste0(round(pct, 1),"%")), size= 2.5, color= "white", hjust= 1 ) + 
  scale_fill_gradient(low= fillLBlue, high=fillBlue) +
  coord_flip()+
  theme_minimal(base_size=8) +
  theme(legend.position="none")

#Relocate
DS %>%
  filter(!is.na(JobRelocateYesNo)) %>%
  group_by(JobRelocateYesNo) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  ggplot(aes(x= reorder(JobRelocateYesNo, count), y = count, fill= count)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = paste0(round(pct, 1),"%")), color="white", hjust=1, size= 2.5) + 
  labs(title= "Willingness to relocate?", x="", y="Count")+
  scale_x_discrete(labels= c("No","Yes"))+
  scale_fill_gradient(low=fillLBlue, high= fillBlue) +
  coord_flip()+
  theme_minimal(base_size= 8) +
  theme(legend.position= "none")

```

## Coding events attended

```{r}
#Code events 
DS_events<- DS%>%
  select(matches("CodeEvent")) %>%
  gather(events, attended, na.rm=TRUE) %>%
  group_by(events) %>%
  summarize(count= n()) %>%
  arrange(desc(count))

DS_events$events<- str_replace(DS_events$events, "CodeEvent", "")

#Code events attended
DS_events %>%
mutate(pct= count/ sum(count) *100) %>%
filter(pct> 0.5) %>%
ggplot(aes(x= reorder(events, count), y= count)) + 
geom_bar(stat= "identity", fill= fillBlue)+
geom_text(aes(label= paste0(round(pct,1),"%")), color= "black", hjust=0, size= 2.3) +
labs(title= "Coding event attended", x= "", y= "Count")+
ylim(c(0, 400)) +
coord_flip() +
theme_minimal(base_size=8) +
theme(legend.position= "none")
```


#Which learning resource do you find usefull

```{r}
#Resources
DS_resource<- DS%>%
  select(matches("Resource")) %>%
  gather(resource, Used, na.rm=TRUE) %>%
  group_by(resource) %>%
  summarize(count= n()) %>%
  arrange(desc(count))

DS_resource$resource<- str_replace(DS_resource$resource, "Resource", "")

#Plot the use of resources
DS_resource %>%
mutate(pct= count/ sum(count) *100) %>%
filter(pct> 0.5) %>%
ggplot(aes(x= reorder(resource, count), y= count)) + 
geom_bar(stat= "identity", fill= fillBlue)+
geom_text(aes(label= paste0(round(pct,1),"%")), color= "black", hjust=0, size= 2.3) +
labs(title= "Learning Resources", x= "", y= "Count")+
ylim(c(0, 500)) +
coord_flip() +
theme_minimal(base_size=8) +
theme(legend.position= "none")

```

## Listening to podcasts

```{r}

#Podcasts
DS_podcast<- DS%>%
  select(matches("Podcast")) %>%
  gather(podcast, Used, na.rm=TRUE) %>%
  group_by(podcast) %>%
  summarize(count= n()) %>%
  arrange(desc(count))

DS_podcast$podcast<- str_replace(DS_podcast$podcast, "Podcast", "")

#Plot
DS_podcast %>%
mutate(pct= count/ sum(count) *100) %>%
ggplot(aes(x= reorder(podcast, count), y= count)) + 
geom_bar(stat= "identity", fill= fillBlue)+
geom_text(aes(label= paste0(round(pct,1),"%")), color= "black", hjust=-0.1, size= 2.3) +
labs(title= "Listening to podcasts", x= "", y= "Count")+
ylim(c(0, 600)) +
coord_flip() +
theme_minimal(base_size=8) +
theme(legend.position= "none")
```

#Demographics

```{r}
#Which country
DS %>%
  filter(!is.na(CountryCitizen)) %>%
  group_by(CountryCitizen) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(10) %>%
  ungroup() %>%
  
  ggplot(aes(x= reorder(CountryCitizen, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
  geom_text(aes(label = paste0(round(pct,1),"%", sep="")),
      position= position_dodge(0), color= "black", size= 2, hjust=-0.1) + 
  ylim(0,300) +
  labs(title= "Country of citizenship", x= "") +
  coord_flip() + theme_minimal(base_size= 8)

#Which language do you speak at home with your family?
DS %>%
  filter(!is.na(LanguageAtHome)) %>%
  group_by(LanguageAtHome) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(10) %>%
  ggplot(aes(x= reorder(LanguageAtHome, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
  geom_text(aes(label = paste0(round(pct,1),"%", sep="")),
      position= position_dodge(0), color= "black", size= 2, hjust= -0.1) + 
  ylim(0, 350) +
  labs(title= "Language at home", x= "") +
  coord_flip() + theme_minimal(base_size= 8)
```

##Educational background

```{r}
#School degree
DS %>%
  filter(!is.na(SchoolDegree)) %>%
  group_by(SchoolDegree) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  ggplot(aes(x= reorder(SchoolDegree, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
  geom_text(aes(label = paste0(round(pct, digits= 2), "%")), hjust= -0.1, size=2) +
  labs(title="School Degree", x="", y= "Count")+
  ylim(c(0,250)) +
  theme_minimal(base_size=8)+
  coord_flip()

#School major
DS %>%
  filter(!is.na(SchoolMajor)) %>%
  group_by(SchoolMajor) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(15) %>%
  ggplot(aes(x= reorder(SchoolMajor, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
geom_text(aes(label = paste0(round(pct, 1),"%")), color= "white", size= 2, hjust= 1) + 
labs(title="School Major", x="", y= "Count")+
coord_flip()+
  theme_minimal(base_size=8)

```

#Employment Status

```{r}

#Changing the null value in the Employment Status variable to "Other" if the respondend provided the other option
emp_status<- within(DS, EmploymentStatus[is.na(EmploymentStatus) & !is.na(EmploymentStatusOther)] <- 'Other')

#Plot
emp_status%>%
 filter(!is.na(EmploymentStatus)) %>%
  group_by(EmploymentStatus) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(15) %>%
  ggplot(aes(x= reorder(EmploymentStatus, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
geom_text(aes(label = paste0(round(pct, 1),"%")), color= "white", size= 2, hjust= 1) + 
labs(title="Employment Status", x="", y= "Count")+
coord_flip()+
  theme_minimal(base_size=8)
  
#Changing the null value in the Employment Field variable to "Other" if the respondend provided the other option
emp_field<- within(DS, EmploymentField[is.na(EmploymentField) & !is.na(EmploymentFieldOther)] <- 'Other')

#Plot
emp_field%>%
 filter(!is.na(EmploymentField)) %>%
  group_by(EmploymentField) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(15) %>%
  ggplot(aes(x= reorder(EmploymentField, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
geom_text(aes(label = paste0(round(pct, 1),"%")), color= "white", size= 2, hjust= 1) + 
labs(title="Employment Field", x="", y= "Count")+
coord_flip()+
  theme_minimal(base_size=8)
  
#Salary
DS %>%
  filter(!is.na(Income)) %>%
  ggplot(aes(x= Income)) + 
  geom_histogram(binwidth = 10000, fill= fillBlue)+
  labs(title= "Current Salary", x= "Income", "Count")+
  theme_minimal(base_size=8)
```


# Age and Income

```{r}
#Correlation
r_gender<- DS%>%
  group_by(Gender) %>%
  filter(Gender %in% c("male","female")) %>%
  summarize(r= cor(Age, Income, use="pairwise.complete.obs"))

#Relations between Age and income
DS %>%
  filter(!is.na(Gender),
        Gender %in% c("female", "male")) %>%
  ggplot(aes(x= Age, y= Income, col = Gender)) + 
  geom_point(alpha= 0.3) +
  facet_wrap(~Gender)+
  geom_smooth(method= 'lm', se=FALSE)+
  geom_text(data= r_gender, mapping= aes(x= -Inf, y= -Inf, label= paste0("r= ", round(r,2))), hjust= -0.1, vjust= -20, color= "black") +
  theme_minimal(base_size=8)
```
