---
title: "Exploring the New Coder Survey 2016"
author: "Muhsina Elampilakkattil"
date: "12 November 2018"
output: 
  html_document:
    toc: true
    fig_width: 6
    fig_height: 3
    code_folding: hide
---
<hr/>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Summary

Free Code Camp is an open source community where you learn to code and build projects for nonprofits.[Free Code Camp](www.freecodecamp.org) and [Code Newbie](www.codenewbie.org) surveyed more than 15,000 people in 2016 who started coding within the past 5 years.

The goal of this survey was to understand these people's motivations in learning to code, how they're learning to code, their demographics, and their socioeconomic background.

<hr/>
#Load Libraries

```{r, message= FALSE, warning= FALSE}
library(tidyverse)
library(RColorBrewer)
library(ggthemes)
library(gridExtra)
library(cowplot)

fillBlue= "#2071a3"
fillLBlue= "#c3d8e5"
fillSilver= "#C0C0C0"

df <- read_csv("Datasets/2016-FCC-New-Coders-Survey-Data.csv")

#Removing the respondents having more than 5 years of coding experience.
survey<- filter(df, MonthsProgramming <= 60)

```

#Who participated?
More than 15000 participated in the survey. Over 1000 participants are having more than **5 years** of programming experience. As I want to analyze the new coders I have filtered out these responses having more than 5 years of experience, considering there are experienced in the field.

* Their average age is **29 years**
* **22%** participants are women
* They have been coding for an average of **13 months**
* **56%** of the respondents live outside the US
* They spend an average of **16 hours** learning each week
* **45%** of them are not native English speakers

```{r, warning= FALSE, fig.width= 10, fig.height= 9}
#Age distribution
plot_age<- survey %>%
  filter(!is.na(Age)) %>%
  ggplot(aes(x= Age)) +
  geom_histogram(binwidth= 2, fill= fillBlue) + 
  labs(title= "Age Distribution", x= "Age", y= "Count", size= 2) + 
  theme_minimal(base_size= 8)  

#Gender Distribution
plot_gender<- survey %>%
  filter(!is.na(Gender)) %>%
  group_by(Gender) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  ungroup() %>%
  ggplot(aes(x= reorder(Gender, count), y= count)) +
  geom_bar(stat= "identity", fill= fillBlue) +
  geom_text(aes(label= paste0(round(pct,2),"%", sep="")),
      position= position_dodge(0), color= "black", size= 2) +
  labs(title= "Gender Distribution", x= "", y= "Count") +
  coord_flip() +
  theme_minimal(base_size= 8)

#How many months have you been programming for?
plot_exp <- survey %>%
  filter(!is.na(MonthsProgramming)) %>%
  ggplot(aes(x= MonthsProgramming))+
  geom_density(aes(y = ..count..), fill = "lightgray", color= fillBlue, alpha= 0.9) + scale_x_log10() +
   labs(title= "Programming experience", x= "Months", y= "Count", size= 2) + 
  theme_minimal(base_size= 8)

#Hours Spend learning each week
plot_learning<- survey %>%
  filter(!is.na(HoursLearning)) %>%
  ggplot(aes(x= HoursLearning))+
  geom_density(aes(y = ..count..), fill = "lightgray", color= fillBlue, alpha= 0.9) + scale_x_log10() +
   labs(title= "Hours learning per week", x= "hours", y= "Count", size= 2) + 
  theme_minimal(base_size= 8)

#Which country are you a citizen of?
plot_country<- survey %>%
  filter(!is.na(CountryCitizen)) %>%
  group_by(CountryCitizen) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(10) %>%
  ungroup() %>%
  
  ggplot(aes(x= reorder(CountryCitizen, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
  geom_text(aes(label = paste0(round(pct,2),"%", sep="")),
      position= position_dodge(0), color= "black", size= 2) + 
  labs(title= "Country of citizenship", x= "") +
  coord_flip() + theme_minimal(base_size= 8)

#Which language do you speak at home with your family?
plot_lang<- survey %>%
  filter(!is.na(LanguageAtHome)) %>%
  group_by(LanguageAtHome) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(10) %>%
  ggplot(aes(x= reorder(LanguageAtHome, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
  geom_text(aes(label = paste0(round(pct,2),"%", sep="")),
      position= position_dodge(0), color= "black", size= 2) + 
  labs(title= "Language at home", x= "") +
  coord_flip() + theme_minimal(base_size= 8)

first_row<- plot_grid(plot_age, plot_gender, align= 'h', rel_widths= c(1,1))
second_row<- plot_grid(plot_exp, plot_country, align= 'h', rel_widths= c(1,1.2))
third_row<- plot_grid(plot_learning, plot_lang, align='h',  rel_widths= c(1,1))
plot_grid(first_row, second_row, third_row, ncol= 1, rel_heights= c(1,1,1))


```

# Which Job Role are you most Interested in?
Majority  of them want to become web developers. Only *quarter* of the respondents opted other than web developer. 

```{r}
#Job roles interested
survey %>%
  filter(!is.na(JobRoleInterest)) %>%
  group_by(JobRoleInterest) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ,
        DS= ifelse(JobRoleInterest=="Data Scientist / Data Engineer", TRUE, FALSE)) %>%
  ggplot(aes(x= reorder(JobRoleInterest, count), y = count, fill= DS)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c(fillSilver, fillBlue)) +
  geom_text(aes(label = paste0(round(pct, digits= 1),"%", sep="")), position= position_dodge(0),hjust= 0, size=2) +
  labs(title= "Job role interest", x="", y="Count") +
  coord_flip() + 
  ylim(c(0,3000)) +
  theme_minimal(base_size= 8)+
  theme(legend.position = "none")
```

# Data Science/ Data Engineers
Lets dive deeper into the details of developing Data scientists or Data engineers.

* 592 of the new coders would like to become *Data Scientists/ Data Engineers*
* **Quarter** of them are women. 
* Their average age is **28 years**

```{r, fig.width= 10}
#Subsetting the data
DS<- filter(survey, JobRoleInterest == "Data Scientist / Data Engineer")

#Age distribution
DS_age<- DS %>%
  filter(!is.na(Age)) %>%
  ggplot(aes(x= Age)) +
  geom_histogram(binwidth= 0.02, fill= fillBlue) + 
  labs(title= "Age Distribution", x= "Age", y= "Count", size= 2) + 
  scale_x_log10() +
  theme_minimal(base_size= 8) 

#Gender Distribution
DS_gender<- DS %>%
  filter(!is.na(Gender)) %>%
  group_by(Gender) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  ungroup() %>%
  ggplot(aes(x= reorder(Gender, count), y= count, fill= count)) +
  geom_bar(stat= "identity") +
  geom_text(aes(label= paste0(round(pct,1),"%", sep="")),
      position= position_dodge(0), hjust= 1.5, color= "white", size= 3) +
  labs(title= "What is your gender?", x= "", y= "Count") +
  scale_fill_gradient(low= fillLBlue, high= fillBlue) +
  coord_flip() +
  theme_minimal(base_size= 8) +
    theme(legend.position = "none")

plot_grid(DS_gender, DS_age, align="h", rel_widths = c(1,1.2))
```

## Programming experience and learning style
581 out of 592 developing Data Scientists/ Engineers responded to the question *how many hours do you spend learning each week?*

* They have been coding for an average of **1 year**
* On an average they spend **14 hours** per week for learning 
```{r, fig.width= 10, warning= FALSE}
#How many months have you been programming for?
DS_exp <- DS %>%
  filter(!is.na(MonthsProgramming)) %>%
  ggplot(aes(x= MonthsProgramming))+
  geom_histogram(binwidth= 0.4, fill = fillBlue) +
  labs(title= "Programming experience", x= "Months", y= "Count", size= 2) + 
  scale_x_continuous(breaks= c(1, 8, 60, 360), labels= c("1", "8", "60", "360"), trans= "log2") +
  theme_minimal(base_size= 8)

#Hours Spend learning each week
DS_learning<- survey %>%
  filter(!is.na(HoursLearning)) %>%
  ggplot(aes(x= HoursLearning))+
  geom_histogram(binwidth= 0.5, fill = fillBlue) +
  labs(title= "Hours learning per week", x= "hours", y= "Count", size= 2) + 
  scale_x_continuous(breaks= c(5,10,20,40,80),labels= c("5","10","20","40","80"), trans="log2") +
  theme_minimal(base_size= 8)

plot_grid(DS_exp, DS_learning, align= "h")
```

##Job Preference

* Majority of them would like to work for a medium sized company
* None of them want to start their on business or do freelance job
* More than half of them would like to work in an office with other developers
* Only **19%** prefer to work remotely. 
* **53%** of them are already applying for jobs or will apply within the next 1 year
* Majority of them are willing to relocate for job

```{r, fig.width=10, fig.height=5}
#Job Preference
p1 <- DS %>%
  group_by(JobPref) %>%
  filter(!is.na(JobPref)) %>%
  summarize(count = n()) %>%
  rbind(data.frame("JobPref"= c("start your own business", "freelance"), count= c(0,0))) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  ggplot(aes(x= reorder(JobPref, count), y = count, fill= count)) + 
  geom_bar(stat = "identity") + 
  scale_x_discrete(drop=FALSE) +
  labs(title="Job preference", x="", y= "Count") +
  geom_text(aes(label = paste0(round(pct, digits= 2),"%")), hjust= 1, color= "white", size= 3) + 
  coord_flip() +
  scale_fill_gradient(low= fillLBlue, high=fillBlue) +
  theme_minimal(base_size= 8) +
  theme(legend.position= "none")

#where would you prefer to work?
p2<- DS %>%
  filter(!is.na(JobWherePref)) %>%
  group_by(JobWherePref) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  ggplot(aes(x= reorder(JobWherePref, count), y = count, fill= count)) + 
  geom_bar(stat = "identity", width= 0.5) + 
  geom_text(aes(label = paste0(round(pct, 1),"%")), color="white", hjust=1, size= 2.5) + 
  labs(title= "Where would you prefere to work?", x="", y="Count")+
  scale_fill_gradient(low=fillLBlue, high= fillBlue) +
  coord_flip()+
  theme_minimal(base_size= 8) +
  theme(legend.position= "none")

#JobApplyWhen
p3<- DS %>%
  filter(!is.na(JobApplyWhen)) %>%
  group_by(JobApplyWhen) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  ggplot(aes(x= reorder(JobApplyWhen, count), y = count, fill= count)) + 
  geom_bar(stat = "identity") + 
  labs(title= "Applying for Jobs?", x="",y="Count") +
  geom_text(aes(label = paste0(round(pct, 1),"%")), size= 2.5, color= "white", hjust= 1 ) + 
  scale_fill_gradient(low= fillLBlue, high=fillBlue) +
  coord_flip()+
  theme_minimal(base_size=8) +
  theme(legend.position="none")

#Relocate
p4<- DS %>%
  filter(!is.na(JobRelocateYesNo)) %>%
  group_by(JobRelocateYesNo) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  ggplot(aes(x= reorder(JobRelocateYesNo, count), y = count, fill= count)) + 
  geom_bar(stat = "identity", width= 0.5) + 
  geom_text(aes(label = paste0(round(pct, 1),"%")), color="white", hjust=1, size= 2.5) + 
  labs(title= "Willingness to relocate?", x="", y="Count")+
  scale_x_discrete(labels= c("No","Yes"))+
  scale_fill_gradient(low=fillLBlue, high= fillBlue) +
  coord_flip()+
  theme_minimal(base_size= 8) +
  theme(legend.position= "none")

r1<- plot_grid(p1, p3, align="h")
r2<- plot_grid(p2, p4, align="h")
plot_grid(r1, r2, ncol= 1, rel_heights = c(1.2,1))
```

## Coding events attended

Most of them have not attended any coding events yet.


```{r}
#Code events 
DS_events<- DS%>%
  select(matches("CodeEvent")) %>%
  gather(events, attended, na.rm=TRUE) %>%
  group_by(events) %>%
  summarize(count= n()) %>%
  arrange(desc(count))

DS_events$events<- str_replace(DS_events$events, "CodeEvent", "")

#Code events attended
DS_events %>%
mutate(pct= count/ sum(count) *100) %>%
filter(pct> 0.5) %>%
ggplot(aes(x= reorder(events, count), y= count)) + 
geom_bar(stat= "identity", fill= fillBlue)+
geom_text(aes(label= paste0(round(pct,1),"%")), color= "black", hjust=0, size= 2.3) +
labs(title= "Coding event attended", x= "", y= "Count")+
ylim(c(0, 400)) +
coord_flip() +
theme_minimal(base_size=8) +
theme(legend.position= "none")
```

##Which learning resource do you find usefull

Half of them are using any of free code camp, codecademy or coursera as learning resource. 

```{r}
#Resources
DS_resource<- DS%>%
  select(matches("Resource")) %>%
  gather(resource, Used, na.rm=TRUE) %>%
  group_by(resource) %>%
  summarize(count= n()) %>%
  arrange(desc(count))

DS_resource$resource<- str_replace(DS_resource$resource, "Resource", "")

#Plot the use of resources
DS_resource %>%
mutate(pct= count/ sum(count) *100) %>%
filter(pct> 0.5) %>%
ggplot(aes(x= reorder(resource, count), y= count)) + 
geom_bar(stat= "identity", fill= fillBlue)+
geom_text(aes(label= paste0(round(pct,1),"%")), color= "black", hjust=0, size= 2.3) +
labs(title= "Learning Resources", x= "", y= "Count")+
ylim(c(0, 500)) +
coord_flip() +
theme_minimal(base_size=8) +
theme(legend.position= "none")

```

## Listening to podcasts

Only **23%** of the developing data scientists listen to coding related podcasts.

```{r}

#Podcasts
DS_podcast<- DS%>%
  select(matches("Podcast")) %>%
  gather(podcast, Used, na.rm=TRUE) %>%
  group_by(podcast) %>%
  summarize(count= n()) %>%
  arrange(desc(count))

DS_podcast$podcast<- str_replace(DS_podcast$podcast, "Podcast", "")

#Plot
DS_podcast %>%
mutate(pct= count/ sum(count) *100) %>%
ggplot(aes(x= reorder(podcast, count), y= count)) + 
geom_bar(stat= "identity", fill= fillBlue)+
geom_text(aes(label= paste0(round(pct,1),"%")), color= "black", hjust=-0.1, size= 2.3) +
labs(title= "Listening to podcasts", x= "", y= "Count")+
ylim(c(0, 600)) +
coord_flip() +
theme_minimal(base_size=8) +
theme(legend.position= "none")
```

## Demographics

* **53%** of them are outside the US
* Nearly half of them are not native English speakers

```{r, fig.width= 10}
#Which country
p1<- DS %>%
  filter(!is.na(CountryCitizen)) %>%
  group_by(CountryCitizen) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(10) %>%
  ungroup() %>%
  
  ggplot(aes(x= reorder(CountryCitizen, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
  geom_text(aes(label = paste0(round(pct,1),"%", sep="")),
      position= position_dodge(0), color= "black", size= 2, hjust=-0.1) + 
  ylim(0,300) +
  labs(title= "Country of citizenship", x= "") +
  coord_flip() + theme_minimal(base_size= 8)

#Which language do you speak at home with your family?
p2<- DS %>%
  filter(!is.na(LanguageAtHome)) %>%
  group_by(LanguageAtHome) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(10) %>%
  ggplot(aes(x= reorder(LanguageAtHome, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
  geom_text(aes(label = paste0(round(pct,1),"%", sep="")),
      position= position_dodge(0), color= "black", size= 2, hjust= -0.1) + 
  ylim(0, 350) +
  labs(title= "Language at home", x= "") +
  coord_flip() + theme_minimal(base_size= 8)
plot_grid(p1,p2,align= "h", rel_widths = c(1.2,1))

```

##Educational background

* **69%** of the developing data scientists have earned atleast a bachelor's degree
* Computer science is the most popular school major

```{r}
#School degree
DS %>%
  filter(!is.na(SchoolDegree)) %>%
  group_by(SchoolDegree) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  ggplot(aes(x= reorder(SchoolDegree, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
  geom_text(aes(label = paste0(round(pct, digits= 2), "%")), hjust= -0.1, size=2) +
  labs(title="School Degree", x="", y= "Count")+
  ylim(c(0,250)) +
  theme_minimal(base_size=8)+
  coord_flip()

#School major
DS %>%
  filter(!is.na(SchoolMajor)) %>%
  group_by(SchoolMajor) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(15) %>%
  ggplot(aes(x= reorder(SchoolMajor, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
geom_text(aes(label = paste0(round(pct, 1),"%")), color= "white", size= 2, hjust= 1) + 
labs(title="School Major", x="", y= "Count")+
coord_flip()+
  theme_minimal(base_size=8)

```

## Employment Status

* Nearly half of them are currently employed
* Around quarter of them are working in software development and IT
* Their average salary is around 48k

```{r, fig.width= 10}

#Changing the null value in the Employment Status variable to "Other" if the respondend provided the other option
emp_status<- within(DS, EmploymentStatus[is.na(EmploymentStatus) & !is.na(EmploymentStatusOther)] <- 'Other')

#Plot
p1<- emp_status%>%
 filter(!is.na(EmploymentStatus)) %>%
  group_by(EmploymentStatus) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(15) %>%
  ggplot(aes(x= reorder(EmploymentStatus, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
geom_text(aes(label = paste0(round(pct, 1),"%")), color= "white", size= 2, hjust= 1) + 
labs(title="Employment Status", x="", y= "Count")+
coord_flip()+
  theme_minimal(base_size=8)
  
#Changing the null value in the Employment Field variable to "Other" if the respondend provided the other option
emp_field<- within(DS, EmploymentField[is.na(EmploymentField) & !is.na(EmploymentFieldOther)] <- 'Other')

#Plot
p2<- emp_field%>%
 filter(!is.na(EmploymentField)) %>%
  group_by(EmploymentField) %>%
  summarize(count = n()) %>%
  mutate(pct = count/ sum(count) * 100 ) %>%
  arrange(desc(count)) %>%
  head(15) %>%
  ggplot(aes(x= reorder(EmploymentField, count), y = count)) + 
  geom_bar(stat = "identity", fill= fillBlue) + 
geom_text(aes(label = paste0(round(pct, 1),"%")), color= "white", size= 2, hjust= 1) + 
labs(title="Employment Field", x="", y= "Count")+
coord_flip()+
  theme_minimal(base_size=8)

plot_grid(p1,p2, align= "h")
```

```{r, fig.width= 5}  
#Salary
DS %>%
  filter(!is.na(Income)) %>%
  ggplot(aes(x= Income)) + 
  geom_histogram(binwidth= 10000, alpha= 0.9, fill= fillBlue) + 
  labs(title= "Current Salary", x= "Income", "Count")+
  theme_minimal(base_size=8)


summary(DS$Income)

```



